Loop Invariant Code Motion in PyPy's Tracing JIT

### 2011-06-13 talk/iwtc11/paper.tex
consistenly use "preamble" and "peeled loop" instead of "first interation" and "second iteration"

\section{Optimizations}
Before the trace is passed to a backend compiling it into machine code
it needs to be optimized to achieve better performance.
The focus of this paper
is loop invariant code motion. The goal of that is to move as many
operations as possible out of the loop making them executed only once
and not every iteration. This we propose to achieve by loop peeling. It
leaves the loop body intact, but prefixes it with one iteration of the
loop. This operation by itself will not achieve anything. But if it is
combined with other optimizations it can increase the effectiveness of
those optimizations. For many optimization of interest some care has
to be taken when they are combined with loop peeling. This is
described below by first explaining the loop peeling optimization
followed by a set of other optimizations and how they interact with
loop peeling.

@@

Loop peeling is achieved by copying the traced iteration of the loop.
The first part (called preamble) finishes with the jump the the second part
(peeled loop). The second part ends up with the jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every other iteration.
The trace from Figure~\ref{fig:unopt-trace} would after this operation become
the trace in Figure~\ref{fig:peeled-trace}.

@@

When applying the following optimizations to this two-iteration trace
some care has to taken as to how the jump arguments of both
iterations and the input arguments of the second iteration are
treated. It has to be ensured that the second iteration stays a proper
trace in the sense that the operations within it only operations on
variables that are either among the input arguments of the second iterations
or are produced within the second iterations. To ensure this we need
to introduce a bit of formalism.

The original trace (prior too peeling) consists of three parts. 
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non-
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the peeled copy, $J$, and jump
arguments $K$. Looking back at our example we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      I &=& \left( p_0, p_1 \right) \\
      J &=& \left( p_0, p_5 \right) \\
      K &=& \left( p_0, p_9 \right) \\
    \end{array}
  %\right.
  .
\end{equation}
To construct the second iteration from the first we also need a
function $m$, mapping the variables of the first iteration onto the
variables of the second. This function is constructed during the
inlining. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means:

@@

Each operation in the trace is inlined in order.
To inline an operation $v=op\left(A_1, A_2, \cdots, A_{|A|}\right)$
a new variable, $\hat v$ is introduced. The inlined operation will
produce $\hat v$ from the input arguments 
\begin{equation}
  \hat v = op\left(m\left(A_1\right), m\left(A_2\right), 
    \cdots, m\left(A_{|A|}\right)\right) . 
\end{equation}
Before the
next operation is inlined, $m$ is extend by assigning $m\left(v\right) = \hat
v$. For the example above, after all the operations have been inlined we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      m\left(p_0\right) &=& p_0 \\
      m\left(p_1\right) &=& p_5 \\
      m\left(i_2\right) &=& i_6 \\
      m\left(i_3\right) &=& i_7 \\
      m\left(i_4\right) &=& i_8 \\
      m\left(p_5\right) &=& p_9 \\
    \end{array}
  %\right.
  .
\end{equation}

@@

No special concerns needs to be taken when implementing redundant
guard removal together with loop peeling. The guards from
the first iteration might make the guards of the second iterations
redundant and thus removed. Therefore the net effect of combining redundant
guard removal with loop peeling is that loop-invariant guards are moved out of the
loop. The second iteration of the example reduces to

@@

The issue at hand is to keep the second iteration a proper
trace. Consider the \lstinline{get} operation on line 19 of
Figure~\ref{fig:unopt-trace}. The result of this operation can be
deduced to be $i_4$ from the \lstinline{set} operation on line
12. Also, the result of the \lstinline{get} operation on line 22 can
be deduced to be $i_3$ from the \lstinline{get} operation on line
8. The optimization will thus remove line 19 and 22 from the trace and
replace $i_6$ with $i_4$ and $i_7$ with $i_3$.

After that, the second
iteration will no longer be in SSA form as it operates on $i_3$ and $i_4$
which are not part of it. The solution is to extend the input
arguments, $J$, with those two variables. This will also extend the
jump arguments of the first iteration, which is also $J$. 
Implicitly that also extends the jump arguments of the second iteration, $K$,
since they are the inlined versions of $J$. For the example $I$ has to
be replaced by $\hat I$ which is formed as a concatenation of $I$ and
$\left(i_3, i_4\right)$. At the same time $K$ has to be replaced by
$\hat K$ which is formed as a concatenation of $K$ and 
$\left(m\left(i_3\right), m\left(i_4\right)\right) = \left(i_7, i_8\right)$. 
The variable $i_7$ will then be replaced by $i_3$ by the heap caching
optimization as it has removed the variable $i_7$. XXX: Maybe we should
replace $i_7=$get(...) with $i_7=i_3$ instead of removing it?

In general what is needed is for the heap optimizer is to keep track of
which variables from the first iterations it reuses in the second
iteration. It has to construct a vector of such variables $H$ which
can be used to update the input and jump arguments,
\begin{equation}
  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H}\right)
\end{equation}
\begin{equation}
  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H})\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized to:

@@

\subsection{Allocation Removals}
By using escape analysis it is possible to identify objects that are
allocated within the loop but never escape it. That is the object are
short lived and no references to them exists outside the loop. This
is performed by processing the operation from top to bottom and
optimistically removing every \lstinline{new} operation. Later on if
it is discovered that a reference to the object escapes the loop, the
\lstinline{new} operation is inserted at this point. All operations
(\lstinline{get} and \lstinline{set}) on the removed objects are also
removed and the optimizer needs to keep track of the value of all
attributes of the object.

Consider again the original unoptimized trace of
Figure~\label{fig:peeled-trace}. Line 10 contains the first
allocation. It is removed and $p_5$ is marked as virtual. This means
that it refers to an virtual object that was not yet
(and might never be) allocated. Line 12 sets the \lstinline{intval}
attribute of $p_5$. This operation is also removed and the optimizer
registers that the attribute \lstinline{intval} of $p_5$ is $i_4$.

When the optimizer reaches line 13 it needs to construct the
arguments for the \lstinline{jump} operation, which contains the virtual
reference $p_5$. This can be achieved by exploding $p_5$ into it's
attributes. In this case there is only one attribute and it's value is
$i_4$, which means the $p_5$ is replaced with $i_4$ in the jump
arguments.

In the general case, each virtual in the jump arguments is exploded into a
vector of variables containing the values of all used attributes. If some
of the attributes are themselves virtuals they are recursively exploded
to make the vector contain only non virtual variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every non
virtual variable of the jump arguments is exploded into a vector. This will
be a vector containing the original variable only. To summarize, for
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is not virtual} \\
        H^{\left(k\right)} & \text{if $J_k$ is virtual}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all non virtual
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
and the arguments of the \lstinline{jump} operation of the second
operation, $K$, are replaced by inlining $\hat J$, 
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

Note that virtuals are only exploded into their attributes when
constructing the arguments of the jump of the first iteration. This
explosion can't be repeated when constructing the arguments of the
jump of the second iteration as it has to mach the first. This means
the objects that was passed as pointers (non virtuals) from the first
iteration to the second also has to be passed as pointers from the
second iteration to the third. If one of these objects are virtual
at the end of the second iteration they need to be allocated right
before the jump. With the simple objects considered in this paper,
that is not a problem. However in more complicated interpreters such
an allocation might, in combination with other optimizations, lead
to additional variables from the first iteration being imported into
the second. This extends both $\hat J$ and $\hat K$, which means that
some care has to be taken, when implementing this, to allow $\hat J$ to
grow while inlining it into $\hat K$.

@@

\subsection{Python}
The python interpreter of the PyPy framework is a complete python
version 2.7 compatible interpreter. A set of numerical
calculations where implemented in both python and in C and their
runtimes compared. The benchmarks are
\begin{itemize}
\item {\bf sqrt}: approximates the square root of $y$ as $x_\infty$
  with $x_0=y/2$ and $x_k = \left( x_{k-1} + y/x_{k-1} \right) /
  2$. There are three different versions of this benchmark where $x_k$
  is represented with different type of objects: int's, float's and
  Fix16's. The later, Fix16, is a custom class that implements
  fixpoint arithmetic with 16 bits precision. In python there is only
  a single implementation of the benchmark that gets specialized
  depending on the class of it's input argument, $y$, while in C,
  there is three different implementations.
\item {\bf conv3}: one dimensional convolution with a kernel of fixed
  size $3$.
\item {\bf conv5}: one dimensional convolution with a kernel of fixed
  size $5$.
\item {\bf conv3x3}: two dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two dimensional
  arrays.
\item {\bf dilate3x3}: two dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the elements, the maximum is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}: an low level video processing algorithm used to
  locate edges in an image. It calculated the gradient magnitude
  using sobel derivatives. The algorithm is in python implemented
  on top of a custom image class that is specially designed for the
  problem. It ensures that there will be no failing guards, and makes
  a lot of the two dimension index calculations loop invariant. The
  intention there is twofold. It shows that the performance impact of
  having wrapper classes giving objects some application specific
  properties is negligible. This is due to the inlining performed
  during the tracing and the allocation removal of the index objects
  introduced. It also shows that it is possible to do some low level
  hand optimizations of the python code and hide those optimization
  under a nice interface without loosing performance.
\end{itemize}

@@

\Subsection{Prolog}
XXX: Carl?

>>>>

\section{Optimizations}
Before the trace is passed to a backend compiling it into machine code
it needs to be optimized to achieve better performance.
The focus of this paper
is loop invariant code motion. The goal of that is to move as many
operations as possible out of the loop making them executed at most once
and not every iteration. This we propose to achieve by loop peeling. It
leaves the loop body intact, but prefixes it with one iteration of the
loop. This operation by itself will not achieve anything. But if it is
combined with other optimizations it can increase the effectiveness of
those optimizations. For many optimization of interest some care has
to be taken when they are combined with loop peeling. This is
described below by first explaining the loop peeling optimization
followed by a set of other optimizations and how they interact with
loop peeling.

@@

Loop peeling is achieved by appending a copy of the traced iteration at
the end of the loop. The copy  is inlined to make the two parts form a
consitant two iteration trace. 
The first part (called preamble) finishes with the jump the the second part
(called peeled loop). The second part ends up with the jump to itself. This way
the preamble will be executed only once while the peeled loop will
be used for every other iteration.
The trace from Figure~\ref{fig:unopt-trace} would after this operation become
the trace in Figure~\ref{fig:peeled-trace}. Line 1-13 shows the
preamble while line 15-27 shows the peeled loop.

@@

When applying the following optimizations to this two-iteration trace
some care has to taken as to how the arguments of the two
\lstinline{jump} operations and the input arguments of the peeled loop are
treated. It has to be ensured that the peeled loop stays a proper
trace in the sense that the operations within it only operates on
variables that are either among its input arguments 
or produced within the peeled loop. To ensure this we need
to introduce a bit of formalism.

The original trace (prior too peeling) consists of three parts. 
A vector of input
variables, $I=\left(I_1, I_2, \cdots, I_{|I|}\right)$, a list of non-
jump operations and a single
jump operation. The jump operation contains a vector of jump variables,
$J=\left(J_1, J_2, \cdots, J_{|J|}\right)$, that are passed as the input variables of the target loop. After
loop peeling there will be a second copy of this trace with input
variables equal to the jump arguments of the pereamble, $J$, and jump
arguments $K$. Looking back at our example we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      I &=& \left( p_0, p_1 \right) \\
      J &=& \left( p_0, p_5 \right) \\
      K &=& \left( p_0, p_9 \right) \\
    \end{array}
  %\right.
  .
\end{equation}
To construct the second iteration from the first we also need a
function $m$, mapping the variables of the preamble onto the
variables of the peeled loop. This function is constructed during the
inlining. It is initialized by mapping the input arguments, $I$, to
the jump arguments $J$,
\begin{equation}
  m\left(I_i\right) = J_i \ \text{for}\ i = 1, 2, \cdots |I| .
\end{equation}
In the example that means:

@@

Each operation in the trace is inlined in order.
To inline an operation $v=\text{op}\left(A_1, A_2, \cdots, A_{|A|}\right)$
a new variable, $\hat v$ is introduced. The inlined operation will
produce $\hat v$ using
\begin{equation}
  \hat v = \text{op}\left(m\left(A_1\right), m\left(A_2\right), 
    \cdots, m\left(A_{|A|}\right)\right) . 
\end{equation}
Before the
next operation is inlined, $m$ is extend by assigning $m\left(v\right) = \hat
v$. For the example above, after all the operations have been inlined we have
\begin{equation}
  %\left\{
    \begin{array}{lcl}
      m\left(p_0\right) &=& p_0 \\
      m\left(p_1\right) &=& p_5 \\
      m\left(i_2\right) &=& i_6 \\
      m\left(i_3\right) &=& i_7 \\
      m\left(i_4\right) &=& i_8 \\
      m\left(p_5\right) &=& p_9 \\
    \end{array}
  %\right.
  .
\end{equation}

@@

No special concerns needs to be taken when implementing redundant
guard removal together with loop peeling. The guards from
the preamble might make the guards of the peeled loop
redundant and thus removed. Therefore the net effect of combining redundant
guard removal with loop peeling is that loop-invariant guards are moved out of the
loop. The peeled loop of the example reduces to

@@

The issue at hand is to keep the peeled loop a proper
trace. Consider the \lstinline{get} operation on line 19 of
Figure~\ref{fig:unopt-trace}. The result of this operation can be
deduced to be $i_4$ from the \lstinline{set} operation on line
12. Also, the result of the \lstinline{get} operation on line 22 can
be deduced to be $i_3$ from the \lstinline{get} operation on line
8. The optimization will thus remove line 19 and 22 from the trace and
replace $i_6$ with $i_4$ and $i_7$ with $i_3$.

After that, the peeled loop
will no longer be in SSA form as it operates on $i_3$ and $i_4$
which are not part of it. The solution is to extend the input
arguments, $J$, with those two variables. This will also extend the
jump arguments of the preamble, which is also $J$. 
Implicitly that also extends the jump arguments of the peeled loop, $K$,
since they are the inlined versions of $J$. For the example $I$ has to
be replaced by $\hat I$ which is formed as a concatenation of $I$ and
$\left(i_3, i_4\right)$. At the same time $K$ has to be replaced by
$\hat K$ which is formed as a concatenation of $K$ and 
$\left(m\left(i_3\right), m\left(i_4\right)\right) = \left(i_7, i_8\right)$. 
The variable $i_7$ will then be replaced by $i_3$ by the heap caching
optimization as it has removed the variable $i_7$. XXX: Maybe we should
replace $i_7=$get(...) with $i_7=i_3$ instead of removing it?

In general what is needed is for the heap optimizer is to keep track of
which variables from the preamble it reuses in the peeled loop.
It has to construct a vector of such variables $H$ which
can be used to update the input and jump arguments,
\begin{equation}
  \hat J = \left(J_1, J_2, \cdots, J_{|J|}, H_1, H_2, \cdots, H_{|H}\right)
\end{equation}
\begin{equation}
  \hat K = \left(K_1, K_2, \cdots, K_{|J|}, m(H_1), m(H_2), \cdots, m(H_{|H})\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized to:

@@

\subsection{Pure operation reusage}
XXX

\subsection{Allocation Removals}
By using escape analysis it is possible to identify objects that are
allocated within the loop but never escape it. That is 
short lived objects with no references outside the loop. This
is performed by processing the operation in order and
optimistically removing every \lstinline{new} operation. Later on if
it is discovered that a reference to the object escapes the loop, the
\lstinline{new} operation is inserted at this point. All operations
(\lstinline{get} and \lstinline{set}) on the removed objects are also
removed and the optimizer needs to keep track of the value of all
used attributes of the object.

Consider again the original unoptimized trace of
Figure~\ref{fig:peeled-trace}. Line 10 contains the first
allocation. It is removed and $p_5$ is marked as virtual. This means
that it refers to an virtual object that was not yet been
(and might never be) allocated. Line 12 sets the \lstinline{intval}
attribute of $p_5$. This operation is also removed and the optimizer
registers that the attribute \lstinline{intval} of $p_5$ is $i_4$.

When the optimizer reaches line 13 it needs to construct the
arguments of the \lstinline{jump} operation, which contains the virtual
reference $p_5$. This can be achieved by exploding $p_5$ into it's
attributes. In this case there is only one attribute and it's value is
$i_4$, which means the $p_5$ is replaced with $i_4$ in the jump
arguments.

In the general case, each virtual in the jump arguments is exploded into a
vector of variables containing the values of all registered attributes. If some
of the attributes are themselves virtuals they are recursively exploded
to make the vector contain only non virtual variables. Some care has
to be taken to always place the attributes in the same order when
performing this explosion. Notation becomes somewhat simpler if also every non
virtual variable of the jump arguments is exploded into a vector. This will
be a vector containing the original variable only. To summarize, for
every variable, $J_k$, of the original jump arguments, $J$, let
\begin{equation}
  \tilde J^{\left(k\right)} = \left\{
      \begin{array}{ll}
        \left(J_k\right)  & \text{if $J_k$ is not virtual} \\
        H^{\left(k\right)} & \text{if $J_k$ is virtual}
      \end{array}
  \right.
  ,
\end{equation}
where $H^{\left(k\right)}$ is a vector containing all non virtual
attributes of $J_k$. The arguments of the optimized \lstinline{jump}
operation are constructed as the concatenation all the $\tilde J^{\left(k\right)}$ vectors,
\begin{equation}
  \hat J = \left( 
    \begin{array}{cccc}
      \tilde J^{\left(1\right)} & \tilde J^{\left(2\right)} & \cdots &
      \tilde J^{\left(|J|\right)} \\
    \end{array}
  \right)      
  .
\end{equation}
and the arguments of the \lstinline{jump} operation of the peeled loop,
$K$, constructed by inlining $\hat J$,
\begin{equation}
  \hat K = \left(m\left(\hat J_1\right), m\left(\hat J_1\right), 
                 \cdots, m\left(\hat J_{|\hat J|}\right)\right)
  .
\end{equation}
In the optimized trace $I$ is replaced by $\hat I$ and $K$ by $\hat
K$. The trace from Figure~\ref{fig:unopt-trace} will be optimized into

@@

Note that virtuals are only exploded into their attributes when
constructing the arguments of the jump of the preamble. This
explosion can't be repeated when constructing the arguments of the
jump of the peeled loop as it has to mach the first. This means
the objects that was passed as pointers (non virtuals) from the first
iteration to the second (from preamble to peeled loop) also has to be
passed as pointers from the second iteration to the third (from peeled
loop to peeled loop). If one of these objects are virtual 
at the end of the peeled loop they need to be allocated right
before the jump. With the simple objects considered in this paper,
that is not a problem. However in more complicated interpreters such
an allocation might, in combination with other optimizations, lead
to additional variables from the first iteration being imported into
the second. This extends both $\hat J$ and $\hat K$, which means that
some care has to be taken, when implementing this, to allow $\hat J$ to
grow while inlining it into $\hat K$. XXX: Maybe we can skip this?

@@

\subsection{Python}
The python interpreter of the PyPy framework is a complete python
version 2.7 compatible interpreter. A set of numerical
calculations where implemented in both python and in C and their
runtimes compared. The benchmarks are
\begin{itemize}
\item {\bf sqrt}: approximates the square root of $y$ as $x_\infty$
  with $x_0=y/2$ and $x_k = \left( x_{k-1} + y/x_{k-1} \right) /
  2$. There are three different versions of this benchmark where $x_k$
  is represented with different type of objects: int's, float's and
  Fix16's. The later, Fix16, is a custom class that implements
  fixpoint arithmetic with 16 bits precision. In python there is only
  a single implementation of the benchmark that gets specialized
  depending on the class of it's input argument, $y$, while in C,
  there are three different implementations.
\item {\bf conv3}: one dimensional convolution with a kernel of fixed
  size $3$.
\item {\bf conv5}: one dimensional convolution with a kernel of fixed
  size $5$.
\item {\bf conv3x3}: two dimensional convolution with kernel of fixed
  size $3 \times 3$ using a custom class to represent two dimensional
  arrays.
\item {\bf dilate3x3}: two dimensional dilation with kernel of fixed
  size $3 \times 3$. This is similar to convolution but instead of
  summing over the elements, the maximum is taken. That places a
  external call to a max function within the loop that prevents some
  of the optimizations.
\item {\bf sobel}: an low level video processing algorithm used to
  locate edges in an image. It calculated the gradient magnitude
  using sobel derivatives. The algorithm is in python implemented
  on top of a custom image class that is specially designed for the
  problem. It ensures that there will be no failing guards, and makes
  a lot of the two dimension index calculations loop invariant. The
  intention here is twofold. It shows that the performance impact of
  having wrapper classes giving objects some application specific
  properties can be negligible. This is due to the inlining performed
  during the tracing and the allocation removal of the index objects
  introduced. It also shows that it is possible to do some low level
  hand optimizations of the python code and hide those optimization
  under a nice interface without loosing performance.
\end{itemize}

@@

\subsection{Prolog}
XXX: Carl?
